---
title: 'Multivariate Analysis: Assignment III Code'
author: "Alexander Ross"
date: "2023-04-03"
output: pdf_document
---

# 1. Load libraries and data set
```{r}
library(corrplot)
library(ggcorrplot)
library(caret) 
library(factoextra)

mnist <- read.csv("emnist-mnist-test.csv")
```
# 2. Exploratory Analysis
### Separate into features and labels
```{r}
mnist_ft = mnist[,-1]
mnist_lb = as.factor(mnist[, 1])
ft_matrix = as.matrix(mnist_ft)
```

### Visualise an example character for each digit
```{r}
ex_digits <- c(5,7,21,19,15,31,14,16,13,35)
for(i in ex_digits){
  vis <- matrix((mnist_ft[i,]), nrow=28, ncol=28)
  vis_num <- apply(vis, 2, as.numeric)
  image(1:28, 1:28, vis_num, col=gray((0:255)/255))
}
```

### Correlations for sample of 9-pixel square in the centre
```{r}
cormat_reduced <- cor(mnist_ft[,c(350:353, 378:381, 406:409)])
corplot_reduced <- corrplot(cormat_reduced)
scatmat_reduced <- ggcorrplot(cormat_reduced)
scatmat_reduced
```

# 3. Dimensionality Reduction
### Perform PCA 
```{r}
pca = prcomp(ft_matrix, center=T)
print(paste('PCs:', ncol(pca$x)))
```

### Visualise labels on PC axes
```{r}
fviz_pca_ind(pca, label="none", habillage=as.factor(mnist_lb))
```

### Proportions of variance visualisations 
```{r}
scree = fviz_eig(pca, ncp=50, addlabels=F, barfill="steelblue")+
  theme(axis.text.x=element_blank(), text=element_text(size=17))
scree
```

```{r}
Sigma <- cov(mnist_ft)
eig <- eigen(Sigma)
prop_var <- eig$values/sum(eig$values)

prop_var_df = data.frame(x=c(1:784), y=cumsum(prop_var))
prop_var_df_50 = prop_var_df[c(1:50),]
prop_var_df_50$colour <- ifelse(prop_var_df_50$y < .8, "red", "black")

ggplot(prop_var_df_50) + geom_point(aes(x, y), col=prop_var_df_50$colour) + 
  geom_hline(yintercept=0.80, linetype="dashed", col="red") +
  annotate("text", x = 10, y = 0.8, label = "80%", vjust = -0.5) +
  xlab("Dimensions") + 
  ylab("Percentage of Explained Variances") + ylim(0,1) +
  theme(text=element_text(size=15))
```

# 4. Clustering 
### Scores of the data in the first 33 dimensions of PCA
```{r}
mnist_ft_scores <- pca$x[,c(1:33)]
```

### Perform clustering with 10 centers
```{r}
set.seed(50)
clustering <- kmeans(x = mnist_ft_scores, centers = 10)
```

### Cluster assignments
```{r}
k_means <- factor(clustering$cluster)
```

### Plot the clusters on PC axis
```{r}
fviz_pca_ind(pca, label = "none", habillage = k_means)
```

### Visualise labels associated with the assignments for each cluster
```{r}
clusters_labels <- data.frame(cbind(k_means, as.factor(mnist_lb)))

ggplot(clusters_labels) + geom_bar(aes(y=mnist_lb, fill=as.factor(k_means))) +
  ggtitle("Cluster Profile for Features Assigned to Each Label") + 
  ylim(0.5,10.5) + scale_y_discrete(limits=factor(c(0:9))) + ylab("Label") + 
  scale_fill_discrete(name = "Cluster")
```
      
## Evaluate clustering performance
### Find the cluster number with the highest frequency among 0s
```{r}
args_0 <- which(mnist_lb %in% 0)
freq0 <- table(k_means[args_0])
most_frequent_0 <- names(which.max(freq0))
```
### among 1s
```{r}
args_1 <- which(mnist_lb %in% 1)
freq1 <- table(k_means[args_1])
most_frequent_1 <- names(which.max(freq1))
```
### among 2s
```{r}
args_2 <- which(mnist_lb %in% 2)
freq2 <- table(k_means[args_2])
most_frequent_2 <- names(which.max(freq2))
```
### among 3s
```{r}
args_3 <- which(mnist_lb %in% 3)
freq3 <- table(k_means[args_3])
most_frequent_3 <- names(which.max(freq3))
```
### among 4s
```{r}
args_4 <- which(mnist_lb %in% 4)
freq4 <- table(k_means[args_4])
most_frequent_4 <- names(which.max(freq4))
```
### among 5s
```{r}
args_5 <- which(mnist_lb %in% 5)
freq5 <- table(k_means[args_5])
most_frequent_5 <- names(which.max(freq5))
```
### among 6s
```{r}
args_6 <- which(mnist_lb %in% 6)
freq6 <- table(k_means[args_6])
most_frequent_6 <- names(which.max(freq6))
```
### among 7s
```{r}
args_7 <- which(mnist_lb %in% 7)
freq7 <- table(k_means[args_7])
most_frequent_7 <- names(which.max(freq7))
```
### among 8s
```{r}
args_8 <- which(mnist_lb %in% 8)
freq8 <- table(k_means[args_8])
most_frequent_8 <- names(which.max(freq8))
```
### among 9s
```{r}
args_9 <- which(mnist_lb %in% 9)
freq9 <- table(k_means[args_9])
most_frequent_9 <- names(which.max(freq9))
```

Since both labels 7 and 9 were assigned cluster 3 most often, we manually 
set 7 to represent cluster 6, the remaining unidientified cluster:
```{r}
most_frequent_7 = "6"
```
### Proportion of correctly clustered characters for each digit
```{r}
accuracy_0 = sum(k_means[args_0]==most_frequent_0)/length(args_0)
accuracy_1 = sum(k_means[args_1]==most_frequent_1)/length(args_1)
accuracy_2 = sum(k_means[args_2]==most_frequent_2)/length(args_2)
accuracy_3 = sum(k_means[args_3]==most_frequent_3)/length(args_3)
accuracy_4 = sum(k_means[args_4]==most_frequent_4)/length(args_4)
accuracy_5 = sum(k_means[args_5]==most_frequent_5)/length(args_5)
accuracy_6 = sum(k_means[args_6]==most_frequent_6)/length(args_6)
accuracy_7 = sum(k_means[args_7]==most_frequent_7)/length(args_7)
accuracy_8 = sum(k_means[args_8]==most_frequent_8)/length(args_8)
accuracy_9 = sum(k_means[args_9]==most_frequent_9)/length(args_9)
accuracy_overall = mean(c(accuracy_0, accuracy_1, accuracy_2, 
                          accuracy_3, accuracy_4, accuracy_5, 
                          accuracy_6, accuracy_7,accuracy_8, accuracy_9))
```

### Calculate the BCSS
```{r}
overall_mean <- colMeans(mnist_ft_scores)
bcss <- sum(clustering$size * apply(clustering$centers, 1, function(x) sum((x - overall_mean)^2)))
```

# 5. Typographical Alteration 
### Add horizontal stroke to a 7
```{r}
first_7 <- matrix((mnist_ft[3,]), nrow=28, ncol=28)
first_7[14,] <- c(rep(0,10),rep(256,14), rep(0,4))
first_7_num <- apply(first_7, 2, as.numeric)
image(1:28, 1:28, first_7_num, col=gray((0:255)/255))
```

### Assign the alteration to all 7s
```{r}
mnist_ft_alt <- mnist_ft

for (i in args_7) {
  ith_7 <- matrix((mnist_ft[i,]), nrow=28, ncol=28)
  ith_7[14,] <- c(rep(0,10),rep(256,14), rep(0,4))
  ith_7_num <- as.vector(apply(ith_7, 2, as.numeric))
  mnist_ft_alt[i,] <- ith_7_num
}
```

### Visualise a 7 to check alteration success
```{r}
check_7 <- matrix((mnist_ft_alt[25,]), nrow=28, ncol=28)
check_7_num <- apply(check_7, 2, as.numeric)
image(1:28, 1:28, check_7_num, col=gray((0:255)/255))
```

# 6. Rerun the analysis for the altered data
```{r}
ft_alt_matrix = as.matrix(mnist_ft_alt)
```

### Perform PCA
```{r}
pca_alt = prcomp(ft_alt_matrix, center=T)
print(paste('PCs:', ncol(pca_alt$x)))
```

### Visualise labels on PC axes
```{r}
fviz_pca_ind(pca_alt, label="none", habillage=as.factor(mnist_lb))
```

### Proportions of variance visualisations
```{r}
scree_alt = fviz_eig(pca_alt, ncp=50, addlabels=T, main='Scree Plot')
scree_alt
```

```{r}
Sigma_alt <- cov(mnist_ft_alt)
eig_alt <- eigen(Sigma_alt)
prop_var_alt <- eig_alt$values/sum(eig_alt$values)

prop_var_alt_df = data.frame(x=c(1:784), y=cumsum(prop_var_alt))
prop_var_alt_df_50 = prop_var_alt_df[c(1:50),]
prop_var_alt_df_50$colour <- ifelse(prop_var_alt_df_50$y < .8, "red", "black")

ggplot(prop_var_alt_df_50) + 
  geom_point(aes(x, y), col=prop_var_alt_df_50$colour) + 
  geom_hline(yintercept=0.80, linetype="dashed", col="red") +
  annotate("text", x = 10, y = 0.8, label = "80%", vjust = -0.5) +
  xlab("Numer of Dimensions") + 
  ylab("Percentage of Explained Variances") + ylim(0,1)
```

## Clustering
### Scores of the data in the first 33 dimensions of PCA
```{r}
mnist_ft_alt_scores <- pca_alt$x[,c(1:33)]
```

### Perform clustering on altered data with 10 centers
```{r}
set.seed(50)
clustering_alt = kmeans(x = mnist_ft_alt_scores, centers = 10)
```

### Cluster assignments
```{r}
k_means_alt = clustering_alt$cluster
```

### Plot the clusters
```{r}
fviz_pca_ind(pca_alt, label = "none", habillage = k_means_alt)
```

## Evaluate clustering performace on altered data

### proportion of correctly clustered characters for each digit
```{r}
accuracy_0_alt = sum(k_means_alt[args_0]==most_frequent_0)/length(args_0)
accuracy_1_alt = sum(k_means_alt[args_1]==most_frequent_1)/length(args_1)
accuracy_2_alt = sum(k_means_alt[args_2]==most_frequent_2)/length(args_2)
accuracy_3_alt = sum(k_means_alt[args_3]==most_frequent_3)/length(args_3)
accuracy_4_alt = sum(k_means_alt[args_4]==most_frequent_4)/length(args_4)
accuracy_5_alt = sum(k_means_alt[args_5]==most_frequent_5)/length(args_5)
accuracy_6_alt = sum(k_means_alt[args_6]==most_frequent_6)/length(args_6)
accuracy_7_alt = sum(k_means_alt[args_7]==most_frequent_7)/length(args_7)
accuracy_8_alt = sum(k_means_alt[args_8]==most_frequent_8)/length(args_8)
accuracy_9_alt = sum(k_means_alt[args_9]==most_frequent_9)/length(args_9)
accuracy_overall_alt = mean(c(accuracy_0_alt, accuracy_1_alt, accuracy_2_alt, 
                          accuracy_3_alt, accuracy_4_alt, accuracy_5_alt, 
                          accuracy_6_alt, accuracy_7_alt,accuracy_8_alt,
                          accuracy_9_alt))
```

### Quantify change in cluster separation
```{r}
overall_mean_alt <- colMeans(mnist_ft_alt_scores)
bcss_alt <- sum(clustering_alt$size * apply(clustering_alt$centers,1,function(x)
                                            sum((x - overall_mean_alt)^2)))
bcss_diff <- bcss_alt-bcss
```